---
title: 'Predicting Daily Bike Traffic Across a Bridge'
author: 'Sean Schilk'
date: "11/8/2025"
output:
  html_document: 
    toc: yes
    toc_float: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    fig_caption: yes
    number_sections: yes
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    theme: lumen
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 5
    fig_height: 4
---

```{css, echo = FALSE}

<style type="text/css">

/* Cascading Style Sheets (CSS) is a stylesheet language used to describe the presentation of a document written in HTML or XML. it is a simple mechanism for adding style (e.g., fonts, colors, spacing) to Web documents. */

h1.title {  /* Title - font specifications of the report title */
  font-size: 24px;
  color: DarkRed;
  font-weight: bold;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}
h4.author { /* Header 4 - font specifications for authors  */
  font-size: 20px;
    font-weight: bold;
  font-family: system-ui;
  color: DarkRed;
  text-align: center;
}
h4.date { /* Header 4 - font specifications for the date  */
  font-size: 18px;
  font-family: system-ui;
    font-weight: bold;
  color: DarkBlue;
  text-align: center;
}
h1 { /* Header 1 - font specifications for level 1 section title  */
    font-size: 22px;
    font-family: system-ui;
    color: navy;
    text-align: left;
      font-weight: bold;
}
h2 { /* Header 2 - font specifications for level 2 section title */
    font-size: 20px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
      font-weight: bold;
}

h3 { /* Header 3 - font specifications of level 3 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
      font-weight: bold;
}

h4 { /* Header 4 - font specifications of level 4 section title  */
    font-size: 18px;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
      font-weight: bold;
}

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

</style>
```
```{r setup, include=FALSE}
# Detect, install and load packages if needed.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
if (!require("mlbench")) {
   install.packages("mlbench")
   library(mlbench)
}
if (!require("MASS")) {
   install.packages("MASS")
   library(MASS)
}
if (!require("readr")) {
   install.packages("readr")
   library(readr)
}
if (!require("psych")) {
   install.packages("psych")
   library(psych)
}
#
# specifications of outputs of code in code chunks
knitr::opts_chunk$set(echo = TRUE,      
                      warning = FALSE,   
                      message = FALSE,  
                      results = TRUE     
                      )   
```

We will be using the same data set as in the previous project, with the same dataset, and mirror the same report. However, this time we will be including a Quasi-Poisson model to the  report.
```{r}
library(openxlsx)
data <- read.xlsx("https://pengdsci.github.io/STA321/ww09/w09-AssignDataSet.xlsx", sheet = paste("data",9, sep = "")) # we got daata set 9
head(data)
```

# EDA

Instead of using two variables HightTemp and LowTemp in the model, we will use the new variable AvgTemp = (HighTemp + LowTemp)/2.

We also discretize Precipitation using the following definition: if Precipitation = 0, then NewPrecip = 0; if Precipitation > 0, then NewPrecip = 1.

```{r}
data$AvgTemp = (data$HighTemp + data$LowTemp)/2

data$NewPrecip <- ifelse(data$Precipitation == 0, 0, 1)
data$NewPrecip <- factor(data$NewPrecip)
```


```{r}
biking <- data # copy for easier notice
biking$Date <- NULL
biking$Day <- as.Date(biking$Day, origin = "1899-12-30")  
biking$Day <- weekdays(biking$Day)
```


We first build a Poisson frequency regression model and ignore the total bike counts for all bridges:
```{r}
library(knitr)
model.freq <- glm(BrooklynBridge ~ Day + AvgTemp + NewPrecip, family = poisson(link = "log"), data = biking)
##
pois.count.coef = summary(model.freq)$coef
kable(pois.count.coef, caption = "The Poisson regression model for the counts of bike crossings versus the predictors")
```
The above inferential table about the regression coefficients indicates significance for every predictor except DayTuesday. Comparing to our baseline DayFriday, each day yields more bike crossings, other than weekdays, where there are more. 

# Poisson Regression on Rates

The following model assesses the potential relationship between bike crossing rates and Day of the week.
```{r}
model.rates <- glm(BrooklynBridge ~ Day + AvgTemp + NewPrecip, 
                   offset = log(Total),
                   family = poisson(link = "log"), 
                   data = biking)
kable(summary(model.rates)$coef, caption = "Poisson regression on the rate of the 
      the bike crossing rate across days of the week")
```
The above table indicates that the log of bike crossing rate is not identical across the days of the week. To be specific, the log rates of the baseline day of Friday were higher on the weekend, but lower than the weekdays. The lowest log rate was Tuesday.

# Quasi - Poisson Rate Model

The above two Poison models assume that there is no dispersion issue in the model. The quasi-Poisson through glm() returns the dispersion coefficient.
```{r}
library(pander)
quasimodel.rates <- glm(BrooklynBridge ~ Day + HighTemp + LowTemp + Precipitation, 
                   offset = log(Total),
                   family = quasipoisson,
                   data = biking)
pander(summary(quasimodel.rates)$coef, caption = "Quasi-Poisson regression on the rate of the bike crossing rate adjusted by day of the week")
```
We use regular Poisson rate model to check whether overdispersion exists. If so, then the quasi-Poisson model is the correct model.
```{r}
ydif=biking$BrooklynBridge-exp(model.rates$linear.predictors)  # diff between y and yhat
prsd = ydif/sqrt(exp(model.rates$linear.predictors))   # Pearson residuals
phi = sum(prsd^2)/15          # Dispersion index: 24-9 = 15  
pander(cbind(Dispersion = phi))
```

We can see dispersion for our Poisson rate model is significantly different than 1, so we will go forward with outr quasi poisson rate model. 

# Final Working Model

The dispersion index is 22.71. It is heavily overdispersed. We go with the Quasi Poisson Rate Regression model. 

The intercept represents the baseline log bike crossing rate on the brooklyn bridge on Friday. The actual rate is exp(-1.878) = 0.1528956 = 15.29%. This is close to the proportion of bikers who crossed the Brooklyn Bridge when compared to the total of bikers crossing all 4 bridges. 

```{r}
biking$newvariable <- biking$BrooklynBridge / biking$Total
hist(biking$newvariable)
mean(biking$newvariable)
```

# Some Visual Comparisons

we can express the quasi rate model as
$$\hat{\text{rate}} = \exp\!\Big(
  -1.878
  - 0.0290 \times \text{DayMonday}
  + 0.0102 \times \text{DaySaturday}
  + 0.0534 \times \text{DaySunday}
  - 0.0299 \times \text{DayThursday}
  - 0.0271 \times \text{DayTuesday}
  - 0.0240 \times \text{DayWednesday}
  + 0.00193 \times \text{HighTemp}
  - 0.00371 \times \text{LowTemp}
  - 0.324 \times \text{Precipitation}
\Big)$$

# Discussions and Conclusions

The regression model based on the bike crossing count is not appropriate since the information on the population size is a key variable in the study of bike crossings distribution. Therefore our Poisson model adjusted for total crossings.

This is only a small data set with limited information. All conclusions in this report are only based on the given data set.


